---
title: "Ecommerce Funnel & Attribution Analysis"
subtitle: "IBM 6530 · Final Presentation"
author: "Bansi Khachar · Matthew Prado · Lucynda Young · Saurabh Parate"
date: today
format:
  revealjs:
    theme: simple
    fontsize: 1.2em
    slide-number: c/t
    progress: true
    transition: fade
    transition-speed: fast
    background-transition: fade
    embed-resources: true
    footer: "IBM 6530 · Ecommerce Attribution Analysis"
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)

funnel_order <- c("page_view", "view_item", "add_to_cart",
                  "begin_checkout", "purchase")

funnel <- data.frame(
  event_name = factor(funnel_order, levels = funnel_order),
  users      = c(2499, 539, 98, 54, 17)
)

table_markov <- data.frame(
  channel = factor(funnel_order, levels = funnel_order),
  percent = c(26.4, 25.9, 21.2, 15.1, 11.4)
)

# Shared chart theme — white background, consistent sizes
theme_slide <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title         = element_text(face = "bold", size = 15,
                                        color = "#1E3A5F",
                                        margin = margin(b = 4)),
      plot.subtitle      = element_text(size = 12.5, color = "#4B5563",
                                        margin = margin(b = 10)),
      axis.text          = element_text(size = 12, color = "#374151"),
      axis.title         = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "#E5E7EB", linewidth = 0.4),
      panel.grid.minor   = element_blank(),
      plot.background    = element_rect(fill = "white", color = NA),
      panel.background   = element_rect(fill = "white", color = NA),
      plot.margin        = margin(8, 14, 6, 14)
    )
}

BLUE <- "#1E3A5F"
ACC  <- "#2563EB"
```

## The Business Problem {background-color="#F8FAFC"}

::: columns
::: {.column width="55%"}
An ecommerce platform is generating strong traffic but converting almost none of it into revenue.

Despite **2,546 distinct users** visiting on a single day, only **17 purchases** were completed.

<br>

::: {style="background:#DBEAFE; border-left:5px solid #1E3A5F; padding:14px 18px; border-radius:0 8px 8px 0;"}
**Overall CVR: 0.7%** — fewer than 1 in 100 visitors completes a purchase
:::
:::

::: {.column width="45%"}
**Why it matters**

- Revenue is being lost at every funnel stage
- Without knowing where users drop off, marketing budget cannot be allocated efficiently
- Without attribution, the business cannot identify which touchpoints drive conversions
:::
:::

---

## Project Objectives {background-color="#F8FAFC"}

::: incremental
1. **Identify and quantify** drop-off points across the purchase funnel using GA4 + BigQuery
2. **Build a stakeholder-ready dashboard** with clean mart logic and correct KPI definitions
3. **Apply attribution analysis** to determine which funnel stages most influence purchase completion
4. **Deliver actionable recommendations** to improve the 0.7% conversion rate
:::

---

## Data & Methodology {background-color="#F8FAFC"}

::: columns
::: {.column width="50%"}
**Data source**

- GA4 Obfuscated Sample
- BigQuery Public Dataset
- Single-day snapshot: Jan 31, 2021
- Accessed via `bigrquery` + `dbplyr` in R

<br>

**Tools**

- R · BigQuery · ChannelAttribution
- Quarto · Looker Studio
:::

::: {.column width="50%"}
**Analytical components**

| Component | Purpose |
|:--|:--|
| Descriptive | KPIs + funnel rates |
| Funnel analysis | Drop-off by stage |
| Markov Chain | Attribution model |
| Heuristic models | First / last touch |
:::
:::

---

## Core KPIs {background-color="#F8FAFC"}

::: columns
::: {.column width="20%"}
:::

::: {.column width="60%"}
| Metric | Value |
|:--|--:|
| Total Events | 26,489 |
| Distinct Users | 2,546 |
| Purchases | **17** |
| Overall CVR | **0.7%** |
:::

::: {.column width="20%"}
:::
:::

<br>

::: {style="background:#DBEAFE; border-left:5px solid #1E3A5F; padding:12px 16px; border-radius:0 8px 8px 0;"}
The platform generates roughly **10 events per user** per session — engagement is not the problem. Conversion is.
:::

---

## Funnel Performance {background-color="#F8FAFC"}

```{r funnel-chart}
#| fig-width: 9
#| fig-height: 4.2
#| fig-alt: "Bar chart of ecommerce funnel. page_view 2499 users, view_item 539, add_to_cart 98, begin_checkout 54, purchase 17. Each stage drops sharply from the previous."

ggplot(funnel, aes(x = event_name, y = users)) +
  geom_col(fill = BLUE, width = 0.6) +
  geom_text(
    aes(label = scales::comma(users)),
    vjust = -0.5, size = 4.2, color = "#1F2937"
  ) +
  scale_x_discrete(
    labels = c("Page View", "View Item", "Add to Cart",
               "Begin Checkout", "Purchase")
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    title    = "Distinct users by funnel stage",
    subtitle = "GA4 Obfuscated Sample — Jan 31, 2021"
  ) +
  theme_slide()
```

---

## Where Users Drop Off {background-color="#F8FAFC"}

```{r dropoff-table}
data.frame(
  Transition = c(
    "page_view → view_item",
    "view_item → add_to_cart",
    "add_to_cart → begin_checkout",
    "begin_checkout → purchase"
  ),
  Dropoff  = c("78.4%", "81.8%", "44.9%", "68.5%"),
  Priority = c("Critical", "Critical", "Moderate", "Critical")
) |>
  kable(
    col.names = c("Transition", "Drop-off", "Priority"),
    align     = c("l", "c", "c")
  ) |>
  kable_styling(
    font_size         = 20,
    full_width        = TRUE,
    bootstrap_options = "basic"
  ) |>
  row_spec(c(1, 2, 4), bold = TRUE,
           color      = "#7F1D1D",
           background = "#FEF2F2") |>
  row_spec(3,
           color      = "#78350F",
           background = "#FFFBEB")
```

::: fragment
::: {style="background:#FEF2F2; border-left:5px solid #991B1B; padding:11px 16px; margin-top:1em; border-radius:0 8px 8px 0;"}
**3 of 4 transitions exceed 68% drop-off.** The product page and checkout experience are the two highest-priority fixes.
:::
:::

---

## Attribution Methodology {background-color="#F8FAFC"}

::: columns
::: {.column width="50%"}
**Setup**

- Unit: individual user session via `user_pseudo_id`
- Touchpoints: all 5 funnel events
- Package: `ChannelAttribution` in R

<br>

**Path defined as:**

`page_view` → `view_item` → `add_to_cart` → `begin_checkout` → `purchase`
:::

::: {.column width="50%"}
**Why Markov Chain?**

| Model | Weakness |
|:--|:--|
| First-touch | Ignores all steps after entry |
| Last-touch | Over-credits the final step |
| **Markov Chain** | Accounts for full journey |

<br>

Markov uses a **removal effect** — it measures how much conversion probability drops when each stage is removed from the path.
:::
:::

---

## Attribution Results {background-color="#F8FAFC"}

```{r markov-chart}
#| fig-width: 8
#| fig-height: 4
#| fig-alt: "Horizontal bar chart of Markov chain attribution. page_view 26.4%, view_item 25.9%, add_to_cart 21.2%, begin_checkout 15.1%, purchase 11.4%."

table_markov |>
  ggplot(aes(x = reorder(channel, percent), y = percent)) +
  geom_col(fill = BLUE, width = 0.6) +
  geom_text(
    aes(label = paste0(percent, "%")),
    hjust = -0.15, size = 4.2, color = "#1F2937"
  ) +
  coord_flip() +
  scale_x_discrete(labels = c(
    "page_view"      = "Page View",
    "view_item"      = "View Item",
    "add_to_cart"    = "Add to Cart",
    "begin_checkout" = "Begin Checkout",
    "purchase"       = "Purchase"
  )) +
  scale_y_continuous(
    limits = c(0, 34),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title    = "Markov chain attribution — % share per stage",
    subtitle = "Credit distributed based on removal effect"
  ) +
  theme_slide()
```

::: fragment
::: {style="background:#DBEAFE; border-left:5px solid #1E3A5F; padding:10px 16px; border-radius:0 8px 8px 0;"}
`page_view` + `view_item` together account for **52.3%** of total conversion credit.
:::
:::

---

## Attribution Model Comparison {background-color="#F8FAFC"}

```{r model-comparison}
#| fig-width: 9
#| fig-height: 4
#| fig-alt: "Stacked bar chart comparing First-Touch, Last-Touch, and Markov attribution. First-touch gives 100% to page_view. Last-touch gives 82% to view_item. Markov distributes evenly across all five stages."

data.frame(
  channel = rep(c("Page View","View Item","Add to Cart",
                  "Begin Checkout","Purchase"), 3),
  model   = rep(c("First-Touch","Last-Touch","Markov"), each = 5),
  value   = c(
    100,  0,    0,    0,    0,
    0,   82.4, 11.8,  5.9,  0,
    26.4, 25.9, 21.2, 15.1, 11.4
  )
) |>
  mutate(
    channel = factor(channel,
                     levels = rev(c("Page View","View Item","Add to Cart",
                                    "Begin Checkout","Purchase"))),
    model   = factor(model, levels = c("First-Touch","Last-Touch","Markov"))
  ) |>
  ggplot(aes(x = model, y = value, fill = channel)) +
  geom_col(position = "fill", width = 0.55) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c(
    "Page View"      = "#1E3A5F",
    "View Item"      = "#2563EB",
    "Add to Cart"    = "#60A5FA",
    "Begin Checkout" = "#93C5FD",
    "Purchase"       = "#BFDBFE"
  )) +
  labs(
    title = "Attribution model comparison — share by funnel stage",
    x     = NULL,
    y     = "Share of attributed conversions",
    fill  = NULL
  ) +
  theme_slide() +
  theme(
    legend.position = "right",
    legend.text     = element_text(size = 12),
    legend.key.size = unit(0.65, "cm")
  )
```

---

## Recommendations {background-color="#F8FAFC"}

::: columns
::: {.column width="33%"}
::: {style="background:#EFF6FF; border-top:4px solid #1E3A5F; border-radius:0 0 8px 8px; padding:14px; font-size:0.95em;"}
**1. Fix landing page**

78.4% of users never view a product

- Featured items above the fold
- Improve search & navigation
- A/B test homepage layout

*Est. +250 users into funnel*
:::
:::

::: {.column width="33%"}
::: {style="background:#EFF6FF; border-top:4px solid #2563EB; border-radius:0 0 8px 8px; padding:14px; font-size:0.95em;"}
**2. Optimize product pages**

81.8% who view a product don't add to cart

- Add reviews & urgency signals
- Benefit-led copy
- A/B test CTA placement

*Est. +54 add-to-cart users*
:::
:::

::: {.column width="33%"}
::: {style="background:#EFF6FF; border-top:4px solid #60A5FA; border-radius:0 0 8px 8px; padding:14px; font-size:0.95em;"}
**3. Reduce checkout friction**

68.5% abandon at checkout

- Enable guest checkout
- Reduce form fields
- Trust badges + mobile UX

*CVR: 0.7% → ~1.1%*
:::
:::
:::

---

## Limitations {background-color="#F8FAFC"}

::: incremental
- **Single-day snapshot** — results cannot account for seasonal variation, day-of-week effects, or campaign cycles. All findings are directional only.
- **Anonymized users** — `user_pseudo_id` does not persist across devices, inflating distinct user counts and deflating true CVR.
- **No channel or campaign data** — cannot separate paid, organic, or direct traffic performance.
- **Small purchase sample** — 17 completed purchases carry high statistical uncertainty. Validate across 30+ days before taking action.
- **Shapley values undefined** — all 17 converting users took identical paths, leaving no variation to compute marginal channel contribution.
:::

---

## Next Steps {background-color="#F8FAFC"}

::: columns
::: {.column width="33%"}
::: {style="background:#EFF6FF; border-top:4px solid #1E3A5F; border-radius:0 0 8px 8px; padding:14px;"}
**Short term (0–30 days)**

- Implement homepage A/B test
- Add reviews to top 10 product pages
- Enable guest checkout
:::
:::

::: {.column width="33%"}
::: {style="background:#EFF6FF; border-top:4px solid #2563EB; border-radius:0 0 8px 8px; padding:14px;"}
**Medium term (30–90 days)**

- Collect 30 days of data for reanalysis
- Implement consistent `user_id` tagging in GA4
- Connect campaign source data to sessions
:::
:::

::: {.column width="33%"}
::: {style="background:#EFF6FF; border-top:4px solid #60A5FA; border-radius:0 0 8px 8px; padding:14px;"}
**Long term (90+ days)**

- Integrate CRM + campaign data into BigQuery
- Run full Markov attribution with 100+ conversions
- Build channel-level ROI model to guide ad spend
:::
:::
:::

---

## Summary {background-color="#F8FAFC"}

::: columns
::: {.column width="55%"}
**What we found**

- CVR of just 0.7% from 2,546 users on a single day
- Largest drop-off: `view_item → add_to_cart` at 81.8%
- Top two funnel stages carry 52% of Markov attribution credit

<br>

**What we recommend**

1. Fix the landing-to-product discovery gap
2. Optimize the product detail page experience
3. Remove checkout friction — CVR can reach ~1.1%
:::

::: {.column width="45%"}
**Deliverables**

- Looker Studio Dashboard
- Quarto Reproducible Report
- This Presentation

<br>

**Data**

- GA4 Obfuscated Sample via BigQuery
- Single-day snapshot: Jan 31, 2021

<br>

:::
:::

---

## Thank You {background-color="#1E3A5F"}

::: {style="text-align:center; padding-top:60px; color:white;"}

**Questions & Discussion**

<br>

::: {style="font-size:0.9em; color:#CBD5E1;"}
Bansi Khachar · Matthew Prado · Lucynda Young · Saurabh Parate

IBM 6530 · Cal Poly Pomona
:::

<br>

::: {style="font-size:0.8em; color:#93C5FD;"}
[github.com/MattPradoCalPoly/project-IBM6530](https://github.com/MattPradoCalPoly/project-IBM6530)
:::

:::
