---
title: "Ecommerce Funnel & Attribution Analysis"
subtitle: "IBM 6530 · Final Presentation"
author: "Bansi Khachar · Matthew Prado · Lucynda Young · Saurabh Parate"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    progress: true
    transition: fade
    transition-speed: fast
    background-transition: fade
    highlight-style: github
    code-fold: true
    embed-resources: true
    css: custom_slides.css
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)

funnel_order <- c("page_view", "view_item", "add_to_cart",
                  "begin_checkout", "purchase")

funnel <- data.frame(
  event_name = factor(funnel_order, levels = funnel_order),
  users      = c(2499, 539, 98, 54, 17),
  cvr        = c("100%", "21.6%", "3.9%", "2.2%", "0.7%"),
  dropoff    = c("—", "78.4%", "81.8%", "44.9%", "68.5%")
)

table_markov <- data.frame(
  channel           = c("page_view","view_item","add_to_cart",
                        "begin_checkout","purchase"),
  total_conversions = c(4, 4, 4, 3, 2),
  percent           = c(26.4, 25.9, 21.2, 15.1, 11.4)
)

theme_slide <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title         = element_text(face = "bold", size = 15,
                                        color = "#111827", margin = margin(b = 6)),
      plot.subtitle      = element_text(size = 12, color = "#6B7280",
                                        margin = margin(b = 10)),
      axis.text          = element_text(size = 11, color = "#6B7280"),
      axis.title         = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "#F3F4F6", linewidth = 0.5),
      panel.grid.minor   = element_blank(),
      plot.background    = element_rect(fill = "white", color = NA),
      panel.background   = element_rect(fill = "white", color = NA),
      plot.margin        = margin(10, 16, 8, 16)
    )
}

C_BLUE   <- "#2563EB"
C_GREEN  <- "#059669"
C_PURPLE <- "#7C3AED"
C_AMBER  <- "#D97706"
C_RED    <- "#DC2626"
```

## The Business Problem {background-color="#1E3A5F"}

::::: columns
::: {.column width="55%"}
### The challenge

An ecommerce platform is generating strong traffic but converting almost none of it into revenue.

Despite **2,546 users** visiting on a single day, only **17 purchases** were completed.

> **Overall CVR: 0.7%**
:::

::: {.column width="45%"}
### Why it matters

-   Revenue is being lost at every funnel stage
-   Without knowing *where* users drop off, budget cannot be allocated efficiently
-   Without attribution, the business cannot identify which touchpoints drive conversions
:::
:::::

------------------------------------------------------------------------

## Project Objectives {background-color="#F8FAFC"}

::: incremental
1.  **Identify and quantify** drop-off points across the ecommerce conversion funnel
2.  **Build a stakeholder-ready dashboard** tracking KPIs with clean BigQuery mart logic
3.  **Apply attribution analysis** to determine which funnel stages have the greatest influence on purchase completion
4.  **Deliver actionable recommendations** to improve the 0.7% conversion rate
:::

------------------------------------------------------------------------

## Data & Methodology {background-color="#F8FAFC"}

::::: columns
::: {.column width="50%"}
**Data Source**

-   GA4 Obfuscated Sample
-   BigQuery Public Dataset
-   Single-day snapshot: Jan 31, 2021
-   Accessed via `bigrquery` + `dbplyr`
:::

::: {.column width="50%"}
**Analytical Framework**

| Component        | Purpose        |
|------------------|----------------|
| Descriptive      | KPIs + funnel  |
| Funnel analysis  | Drop-off rates |
| Markov Chain     | Attribution    |
| Heuristic models | Comparison     |
:::
:::::

------------------------------------------------------------------------

## Core KPIs {background-color="#F8FAFC"}

:::::: columns
::: {.column width="20%"}
:::

::: {.column width="60%"}
| Metric          |    Value |
|:----------------|---------:|
| Total Events    |   26,489 |
| Distinct Users  |    2,546 |
| Purchases       |       17 |
| **Overall CVR** | **0.7%** |
:::

::: {.column width="20%"}
:::
::::::

::: fragment
> The platform generates **\~10 events per user per session** — engagement is not the problem. Conversion is.
:::

------------------------------------------------------------------------

## Funnel Performance {background-color="#F8FAFC"}

```{r funnel-chart, fig.width=9, fig.height=4.5}
ggplot(funnel, aes(x = event_name, y = users)) +
  geom_col(fill = C_BLUE, width = 0.6) +
  geom_text(aes(label = scales::comma(users)),
            vjust = -0.5, size = 4, color = "#374151") +
  scale_x_discrete(labels = c("Page View","View Item",
                               "Add to Cart","Begin Checkout","Purchase")) +
  scale_y_continuous(labels = scales::comma,
                     expand  = expansion(mult = c(0, 0.15))) +
  labs(title    = "Distinct users by funnel stage",
       subtitle = "GA4 Obfuscated Sample | Jan 31, 2021") +
  theme_slide()
```

------------------------------------------------------------------------

## Where Users Drop Off {background-color="#F8FAFC"}

```{r dropoff-table}
data.frame(
  Stage     = c("page_view → view_item",
                "view_item → add_to_cart",
                "add_to_cart → begin_checkout",
                "begin_checkout → purchase"),
  `Drop-off` = c("78.4%", "81.8%", "44.9%", "68.5%"),
  Priority   = c("Critical", "Critical", "Moderate", "Critical")
) |>
  kable(align = c("l","c","c"), col.names = c("Transition","Drop-off","Priority")) |>
  kable_styling(font_size = 18, full_width = TRUE) |>
  row_spec(c(1,2,4), color = "#a32d2d", bold = TRUE) |>
  row_spec(3, color = "#b45309")
```

:::: fragment
::: callout-important
**3 of 4 transitions have drop-off rates above 68%**. The product detail page and checkout experience are the two highest-priority fixes.
:::
::::

------------------------------------------------------------------------

## Attribution Methodology {background-color="#F8FAFC"}

::::: columns
::: {.column width="50%"}
**Approach**

-   Unit: individual user session via `user_pseudo_id`
-   Touchpoints: all 5 funnel events
-   Model: **Markov Chain (Order 1)**
-   Comparison: First-touch & Last-touch heuristics
:::

::: {.column width="50%"}
**Why Markov Chain?**

| Model       | Limitation                     |
|-------------|--------------------------------|
| First-touch | Ignores everything after entry |
| Last-touch  | Over-credits final step        |
| **Markov**  | Accounts for full journey      |
:::
:::::

------------------------------------------------------------------------

## Attribution Results {background-color="#F8FAFC"}

```{r markov-chart, fig.width=8, fig.height=4.2}
table_markov |>
  mutate(channel = factor(channel, levels = funnel_order)) |>
  arrange(channel) |>
  ggplot(aes(x = reorder(channel, percent), y = percent)) +
  geom_col(fill = C_BLUE, width = 0.6) +
  geom_text(aes(label = paste0(percent, "%")),
            hjust = -0.1, size = 4, color = "#374151") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 35),
                     expand  = expansion(mult = c(0, 0.05))) +
  scale_x_discrete(labels = c("purchase"       = "Purchase",
                               "begin_checkout" = "Begin Checkout",
                               "add_to_cart"    = "Add to Cart",
                               "view_item"      = "View Item",
                               "page_view"      = "Page View")) +
  labs(title    = "Markov chain attribution — % share per stage",
       subtitle = "Credit distributed based on removal effect") +
  theme_slide()
```

::: fragment
`page_view` + `view_item` account for **52.3%** of total conversion credit — the top of the funnel matters most.
:::

------------------------------------------------------------------------

## Model Comparison {background-color="#F8FAFC"}

```{r model-comparison-chart, fig.width=9, fig.height=4.2}
comparison <- data.frame(
  channel  = rep(c("Page View","View Item","Add to Cart",
                   "Begin Checkout","Purchase"), 3),
  model    = rep(c("First-Touch","Last-Touch","Markov"), each = 5),
  value    = c(100, 0, 0, 0, 0,
               0, 82.4, 11.8, 5.9, 0,
               26.4, 25.9, 21.2, 15.1, 11.4)
) |>
  mutate(
    channel = factor(channel,
                     levels = c("Page View","View Item","Add to Cart",
                                "Begin Checkout","Purchase")),
    model   = factor(model, levels = c("First-Touch","Last-Touch","Markov"))
  )

ggplot(comparison, aes(x = model, y = value, fill = channel)) +
  geom_col(position = "fill", width = 0.55) +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = c(
    "Page View"      = "#2563EB",
    "View Item"      = "#059669",
    "Add to Cart"    = "#7C3AED",
    "Begin Checkout" = "#D97706",
    "Purchase"       = "#9CA3AF"
  )) +
  labs(title = "Attribution model comparison",
       x = NULL, y = "Share", fill = NULL) +
  theme_slide() +
  theme(legend.position = "right",
        legend.text = element_text(size = 11))
```

------------------------------------------------------------------------

## Recommendations {background-color="#F8FAFC"}

:::::: columns
::: {.column width="33%"}
**1. Fix landing page**

78.4% of users never view a product

-   Featured products above the fold
-   Improve search & navigation
-   A/B test homepage layout

*+250 users into funnel*
:::

::: {.column width="33%"}
**2. Optimize product pages**

81.8% who view a product don't add to cart

-   Add reviews & urgency signals
-   Rewrite benefit-led copy
-   A/B test CTA placement

*+54 add-to-cart users*
:::

::: {.column width="33%"}
**3. Reduce checkout friction**

68.5% abandon at checkout

-   Guest checkout
-   Fewer form fields
-   Trust badges + mobile optimization

*CVR: 0.7% → 1.1%*
:::
::::::

------------------------------------------------------------------------

## Limitations & Caveats {background-color="#FFF7ED"}

::: incremental
-   **Single-day snapshot** — findings cannot account for seasonal variation, day-of-week effects, or campaign cycles. All results are directional.
-   **Anonymized users** — `user_pseudo_id` does not persist across devices. Distinct user count is inflated; true CVR is likely higher than 0.7%.
-   **No channel data** — cannot determine whether paid, organic, or direct traffic converts differently.
-   **Small sample** — 17 purchases carries high statistical uncertainty. Validate across 30+ days before acting.
-   **Shapley values undefined** — all converting users took identical paths, leaving no variation to calculate marginal contribution.
:::

------------------------------------------------------------------------

## Next Steps {background-color="#F0FDF4"}

::::: columns
::: {.column width="50%"}
**Short term (0–30 days)**

-   Implement homepage A/B test
-   Add reviews to top 10 product pages
-   Enable guest checkout

**Medium term (30–90 days)**

-   Collect 30 days of data for reanalysis
-   Implement consistent `user_id` tagging
:::

::: {.column width="50%"}
**Long term (90+ days)**

-   Integrate CRM + campaign data into BigQuery
-   Run full multi-touch Markov attribution with 100+ conversions
-   Build channel-level ROI model to guide budget allocation
:::
:::::

------------------------------------------------------------------------

## Summary {background-color="#1E3A5F"}

::::: columns
::: {.column width="60%"}
**What we found**

-   CVR of just 0.7% from 2,546 users
-   Largest drop: `view_item → add_to_cart` at 81.8%
-   Top of funnel carries 52% of Markov attribution credit

**What we recommend**

1.  Fix the landing-to-product discovery drop
2.  Optimize the product detail page
3.  Remove checkout friction
:::

::: {.column width="40%"}
**Deliverables**

-   Looker Studio Dashboard
-   Quarto Reproducible Report
-   This Presentation

**Repository**

[github.com/MattPradoCalPoly/project-IBM6530](https://github.com/MattPradoCalPoly/project-IBM6530)
:::
:::::

------------------------------------------------------------------------

##  {background-color="#1E3A5F"}

::: {style="text-align: center; padding-top: 80px;"}
# Thank You

**Questions?**

Bansi Khachar · Matthew Prado · Lucynda Young · Saurabh Parate

*IBM 6530 · Cal Poly Pomona*
:::
