---
title: "Ecommerce Funnel & Attribution Analysis"
subtitle: "IBM 6530 · Deliverable 2 — Final Quarto Report"
author: "Bansi Khachar, Matthew Prado, Lucynda Young, Saurabh Parate"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    smooth-scroll: true
execute:
  echo: true
  warning: false
  message: false
---

---

## Narrative Summary of the Business Problem

The client operates an ecommerce platform struggling to convert site visitors into paying customers. On January 31, 2021, the platform generated **26,489 events** from **2,546 distinct users**, yet only **17 purchases** were completed — a conversion rate of just **0.7%**.

The data reveals that users are actively engaging with the platform but are abandoning the funnel at multiple stages before reaching purchase. The core business problem is not a lack of traffic; it is a broken conversion architecture that is leaking potential revenue at every stage of the funnel, with the most severe losses occurring at the product discovery stage.

**Why it matters:**

- High traffic with low conversion signals a disconnect between user intent and purchase completion
- Unidentified drop-off points make it impossible to allocate marketing spend efficiently
- Without attribution data, the business cannot determine which funnel stages drive the most conversions

---

## Setup & Data Connection

```{r setup}
library(DBI)
library(bigrquery)
library(dplyr)
library(dbplyr)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)
library(ChannelAttribution)
library(tidyr)
library(htmltools)

bigrquery::bq_auth(
  scopes = "https://www.googleapis.com/auth/bigquery"
)

project_id <- "ga4-analysis-491223"

mart_project <- "bigquery-public-data"
mart_dataset <- "ga4_obfuscated_sample_ecommerce"
mart_view <- "events_20210131"

con <- DBI::dbConnect(
  bigrquery::bigquery(),
  project = mart_project,
  dataset = mart_dataset,
  billing = project_id
)

mart_tbl <- dplyr::tbl(con, mart_view)
```

---

## Descriptive Insights

### Core KPIs

```{r kpis}
kpis <- mart_tbl |>
  summarise(
    total_events   = n(),
    distinct_users = n_distinct(user_pseudo_id),
    purchases      = sum(as.integer(event_name == "purchase"), na.rm = TRUE)
  ) |>
  collect() |>
  mutate(cvr = paste0(round(purchases / distinct_users * 100, 1), "%"))

kpis |>
  kable(
    col.names = c("Total Events", "Distinct Users", "Purchases", "CVR"),
    align     = "c",
    caption   = "Core KPIs — GA4 Obfuscated Sample, Jan 31 2021"
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

**Interpretation:** The platform recorded 26,489 total events from 2,546 distinct users but completed only 17 purchases, giving an overall conversion rate of 0.7%. Fewer than 1 in 100 users who visited the site ended up buying something.

---

### Funnel Performance

```{r funnel-table}
funnel_order <- c("page_view", "view_item", "add_to_cart",
                  "begin_checkout", "purchase")

funnel <- mart_tbl |>
  filter(event_name %in% funnel_order) |>
  group_by(event_name) |>
  summarise(users = n_distinct(user_pseudo_id)) |>
  collect() |>
  mutate(event_name = factor(event_name, levels = funnel_order)) |>
  arrange(event_name) |>
  mutate(
    cvr     = paste0(round(users / first(users) * 100, 1), "%"),
    dropoff = paste0(round(
      (1 - users / lag(users, default = first(users))) * 100, 1
    ), "%")
  )

funnel |>
  kable(
    col.names = c("Event", "Users", "CVR", "Drop-off"),
    align     = "c",
    caption   = "Funnel Performance — Distinct Users by Stage"
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |>
  row_spec(2, color = "#a32d2d") |>
  row_spec(3, color = "#a32d2d") |>
  row_spec(5, color = "#a32d2d")
```

**Key findings:**

- The steepest drop-off occurs between `page_view` and `view_item` at **78.4%** — nearly 4 in 5 users who landed on the site never engaged with a product
- A second critical drop-off of **81.8%** occurs between `view_item` and `add_to_cart` — users who found a product still overwhelmingly chose not to act on it
- The final drop between `begin_checkout` and `purchase` is **68.5%** — more than two thirds of users with clear purchase intent abandoned at the last step

---

### Funnel Bar Chart

```{r funnel-chart, fig.width=8, fig.height=4}
ggplot(funnel, aes(x = event_name, y = users)) +
  geom_col(fill = "#2563EB", width = 0.6) +
  geom_text(
    aes(label = scales::comma(users)),
    vjust = -0.5, size = 3.5, color = "#374151"
  ) +
  scale_x_discrete(
    labels = c("Page View", "View Item", "Add to Cart",
               "Begin Checkout", "Purchase")
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Ecommerce Funnel — Distinct Users by Stage",
    subtitle = "GA4 Obfuscated Sample | Jan 31, 2021",
    x        = NULL,
    y        = "Distinct Users"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "#6B7280", size = 10),
    axis.text     = element_text(color = "#6B7280"),
    panel.grid.major.x = element_blank()
  )
```

**Interpretation:** The bar chart makes the funnel collapse immediately visible. The dramatic drop from 2,499 page views to 539 product views is the single biggest opportunity — most users are bouncing before they ever see a product.

---

## Advanced Analysis — Attribution

### Methodology

```{r attribution-build}
# Step 1 — pull purchaser IDs into local memory
purchaser_ids <- mart_tbl |>
  filter(event_name == "purchase") |>
  select(user_pseudo_id) |>
  distinct() |>
  collect() |>
  pull(user_pseudo_id)

# Step 2 — pull all funnel events for converting users
purchasers <- mart_tbl |>
  filter(
    user_pseudo_id %in% purchaser_ids,
    event_name     %in% funnel_order
  ) |>
  select(user_pseudo_id, event_name) |>
  collect()

# Step 3 — build channel_stack for ChannelAttribution
channel_stack <- purchasers |>
  group_by(user_pseudo_id) |>
  summarise(
    cleaned_path = paste(event_name, collapse = ">"),
    .groups      = "drop"
  ) |>
  mutate(
    conversion     = 1,
    non_conversion = 0
  ) |>
  group_by(cleaned_path) |>
  summarise(
    conversion     = sum(conversion),
    non_conversion = sum(non_conversion),
    .groups        = "drop"
  )
```

**Unit of analysis:** Individual user session tracked via `user_pseudo_id`, defined as any event path from `page_view` to `purchase`.

**Touchpoints:** `page_view` → `view_item` → `add_to_cart` → `begin_checkout` → `purchase`

**Model used:** Markov Chain Attribution (Order 1) — models transition probabilities between each funnel stage and estimates each stage's contribution using a removal effect. More reliable than first-touch or last-touch because it accounts for the full journey.

---

### First-Touch & Last-Touch Attribution

```{r heuristic}
heuristic_results <- heuristic_models(
  channel_stack,
  var_path = "cleaned_path",
  var_conv = "conversion",
  sep      = ">"
)

heuristic_results |>
  select(channel_name, first_touch, last_touch) |>
  kable(
    col.names = c("Funnel Stage", "First-Touch", "Last-Touch"),
    digits    = 3,
    align     = "c",
    caption   = "Heuristic Attribution — Conversion Counts by Model"
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

**Interpretation:** First-touch assigns all 17 conversions to `page_view` because every buyer started there. Last-touch gives 14 of 17 to `view_item` because that was typically the final step. Neither model tells the full story — one only sees entry, the other only sees exit.

---

### Markov Chain Attribution

```{r markov}
markov <- markov_model(
  channel_stack,
  var_path = "cleaned_path",
  var_conv = "conversion",
  var_null = "non_conversion",
  order    = 1,
  sep      = ">"
)

table_markov <- data.frame(
  channel           = markov$channel_name,
  total_conversions = round(markov$total_conversions),
  percent           = round(
    markov$total_conversions / sum(channel_stack$conversion) * 100, 1
  )
)

table_markov |>
  kable(
    col.names = c("Funnel Stage", "Attributed Conversions", "% Share"),
    align     = "c",
    caption   = "Markov Chain Attribution Results"
  ) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

**Interpretation:** The Markov model distributes credit relatively evenly across all five stages, confirming that every touchpoint plays a meaningful role. `page_view` (26.4%) and `view_item` (25.9%) carry the highest combined weight at over 52% of total attributed credit, reinforcing that the top of the funnel is where the most conversion influence lies.

---

### Model Comparison

```{r model-comparison}
model_comparison <- heuristic_results |>
  select(channel_name, first_touch, last_touch) |>
  left_join(
    table_markov |> select(channel, percent),
    by = c("channel_name" = "channel")
  ) |>
  rename(
    `Funnel Stage` = channel_name,
    `First-Touch`  = first_touch,
    `Last-Touch`   = last_touch,
    `Markov (%)`   = percent
  )

model_comparison |>
  kable(digits = 3, align = "c",
        caption = "Attribution Model Comparison") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

---

### Attribution Model Comparison Chart

```{r model-chart, fig.width=8, fig.height=4}
plot_data <- heuristic_results |>
  select(channel_name, first_touch, last_touch) |>
  left_join(
    table_markov |>
      mutate(markov = percent / 100) |>
      select(channel, markov),
    by = c("channel_name" = "channel")
  ) |>
  tidyr::pivot_longer(
    cols      = c(first_touch, last_touch, markov),
    names_to  = "model",
    values_to = "value"
  ) |>
  mutate(
    model        = dplyr::recode(model,
      first_touch = "First-Touch",
      last_touch  = "Last-Touch",
      markov      = "Markov Chain"
    ),
    channel_name = factor(channel_name, levels = funnel_order)
  )

ggplot(plot_data, aes(x = model, y = value, fill = channel_name)) +
  geom_col(position = "fill", width = 0.55) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c(
    "page_view"      = "#2563EB",
    "view_item"      = "#059669",
    "add_to_cart"    = "#7C3AED",
    "begin_checkout" = "#D97706",
    "purchase"       = "#6B7280"
  )) +
  labs(
    title = "Attribution model comparison — share by funnel stage",
    x     = "Attribution Model",
    y     = "Share of Attributed Conversions",
    fill  = "Funnel Stage"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold", size = 13),
    panel.grid.major.x = element_blank(),
    axis.text          = element_text(color = "#6B7280")
  )
```

---

### Transition Matrix Heatmap

```{r transition-heatmap, fig.width=7, fig.height=5}
trans_1st <- transition_matrix(
  channel_stack,
  var_path = "cleaned_path",
  var_conv = "conversion",
  var_null = "non_conversion",
  order    = 1,
  sep      = ">"
)

ggplot(
  trans_1st$transition_matrix,
  aes(x = channel_to, y = channel_from, fill = transition_probability)
) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(
    aes(label = round(transition_probability, 2)),
    size = 4, color = "#1a1a1a"
  ) +
  scale_fill_gradient(low = "#DBEAFE", high = "#1D4ED8") +
  labs(
    title = "Funnel transition matrix",
    x     = "Next stage",
    y     = "Previous stage",
    fill  = "Transition\nProbability"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold", size = 13),
    axis.text.x  = element_text(angle = 30, hjust = 1, color = "#6B7280"),
    axis.text.y  = element_text(color = "#6B7280"),
    panel.grid   = element_blank()
  )
```

**Interpretation:** The heatmap shows the probability of a user moving from one funnel stage to the next. Darker blue cells indicate stronger transitions. Any light-colored cells on the conversion diagonal show where transition probability is weak — those are your highest-priority friction points.

---

## Recommendations

### Recommendation 1 — Fix Landing-to-Product Discovery Drop-off

The **78.4% drop** between `page_view` and `view_item` is the highest-priority problem. Of 2,499 users who landed on the site, only 539 ever viewed a product. Most users are bouncing before they engage with any product at all.

**Actions:**

- Add prominent product categories and featured items above the fold
- Improve internal search and filtering so users find relevant products quickly
- A/B test homepage layouts to identify which entry points drive the most `view_item` events

**Expected impact:** A 10-point improvement adds approximately 250 more users into the product funnel, more than doubling the downstream conversion pool.

---

### Recommendation 2 — Optimize the Product Detail Page

Of the 539 users who viewed a product, only 98 added one to their cart — an **81.8% drop-off**. Users are reaching the product but not acting on it.

**Actions:**

- Add customer reviews and star ratings to all product pages
- Introduce urgency signals such as low stock indicators or limited-time offers
- Rewrite product descriptions to lead with benefits rather than features
- A/B test CTA button copy, size, and placement

**Expected impact:** A 10-point improvement increases add-to-cart users from 98 to approximately 152, directly expanding the checkout pool.

---

### Recommendation 3 — Reduce Checkout Abandonment

Of the 54 users who began checkout, only 17 completed a purchase — a **68.5% abandonment rate**. These users had clear purchase intent, so the barrier is the checkout process itself.

**Actions:**

- Implement guest checkout to remove account creation friction
- Reduce required form fields to the minimum necessary
- Display security badges, accepted payment methods, and return policy prominently
- Ensure checkout is fully optimized for mobile

**Expected impact:** Reducing abandonment by 20 points would increase purchases from 17 to approximately 27, lifting CVR from 0.7% to approximately 1.1%.

---

## Limitations

| Limitation | Detail |
|---|---|
| **Single-day snapshot** | Covers only Jan 31, 2021. Cannot account for seasonal variation, day-of-week effects, or campaign cycles. All findings are directional. |
| **Anonymized users** | `user_pseudo_id` does not persist across devices. One real user on mobile and desktop counts as two, inflating distinct user counts and deflating true CVR. |
| **No channel data** | Without source and medium attribution, it is impossible to determine whether paid, organic, direct, or referral traffic converts differently. |
| **Small purchase sample** | 17 purchases carry high statistical uncertainty. Results should not be treated as definitive without replication across a broader time window. |
| **Shapley values undefined** | All 17 converting users passed through identical funnel stages, leaving no variation to calculate marginal contribution. A minimum of 50–100 conversions with varied paths is needed. |

---

## Reproducible Code & Instructions

### Requirements

Install the following R packages before rendering:

```r
install.packages(c(
  "bigrquery", "dplyr", "dbplyr", "ggplot2",
  "scales", "knitr", "kableExtra",
  "ChannelAttribution", "tidyr", "DBI", "htmltools"
))
```

### Authentication

```r
bigrquery::bq_auth(
  scopes = "https://www.googleapis.com/auth/bigquery"
)
```

Authenticate with the Google account that has BigQuery read access on the `bigquery-489601` project.

### Render

From your terminal in the project directory:

```bash
quarto render Deliverable_2_Final.qmd
```

Or from RStudio: open the file and click **Render**.

### Repository

All code, data connections, and rendered outputs are available at:
[https://github.com/MattPradoCalPoly/project-IBM6530](https://github.com/MattPradoCalPoly/project-IBM6530)

```{r close}
DBI::dbDisconnect(con)
```
